---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
***
  
# Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement -- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.<br><br><br><br>

# Data

The data for this assignment can be downloaded from the course web site:  
  
* Dataset: <a href="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip">Activity monitoring data</a> [52K]

The variables included in this dataset are:  

* __steps__: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* __date__: The date on which the measurement was taken in YYYY-MM-DD format
* __interval__: Identifier for the 5-minute interval in which measurement was taken
  
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
<br><br><br><br>

# Performed analysis
In this section you will find the description of the questions and the explanations and the code used in order to answer them.
<br><br>

## Environment settings
My first step was to change the locale into English and load some libraries that I found I was going to need. I setted the working directory too, but I found it was pointless to add the code for my directory. If you are going to execute this in your computer, just remember to set the working directory pointing where you have downloaded the data.  

```{r "Presettings", message = FALSE, warning = FALSE, results = 'hide'}
Sys.setlocale("LC_ALL","en_GB.UTF-8");
library("dplyr");
```
<br><br>

## Loading and preprocessing the data
  
1. __Load the data__
  
2. __Process/transform the data (if necessary) into a format suitable for your analysis__
  
I loaded the data into a format suitable for the analysis. The original file is unzipped and loaded into _activityData_ variable.  

```{r}
untar('activity.zip');
activityData <- read.csv('activity.csv', header = TRUE, sep = ",");
```
<br><br>

## What is mean total number of steps taken per day?

1. __Calculate the total number of steps taken per day.__  

I use the _aggregate()_ function to sum the steps, grouping by date. The result is saved into _sumStepsByDate_. Then, I change the names of the columns of this variable into a more meaningful ones.  

```{r}
sumStepsByDate <- aggregate(activityData$steps, by=activityData["date"], FUN=sum);
names(sumStepsByDate) <- c("date","sumsteps");
```
  
<br>
2. __Make a histogram of the total number of steps taken each day__  

This is performed using the function _hist()_ over the sum of the steps.  

```{r}
hist(sumStepsByDate$sumsteps, main="Histogram of the total number of steps by day", 
     xlab="Total steps by day");
```
  
<br>
3. __Calculate and report the mean and median of the total number of steps taken per day__  

Using _mean()_ and _median()_ functions.  
    
```{r}
mean(sumStepsByDate$sumsteps, na.rm = TRUE);
median(sumStepsByDate$sumsteps, na.rm = TRUE);
```
<br><br>

## What is the average daily activity pattern?
    
1. __Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)__  

I use the _aggregate()_ function to calculate the mean of the steps, grouping by interval. The result is saved into _avgStepsByInterval_. Then, I change the names of the columns of this variable into a meaningful ones. Third, I use the Base plotting system plot function to show the mean of steps taken in each 5-minute interval.  

```{r}
avgStepsByInterval <- aggregate(activityData$steps, by=activityData["interval"], FUN=mean, na.rm = TRUE);
names(avgStepsByInterval) <- c("interval","avgsteps");
plot(avgStepsByInterval$interval, avgStepsByInterval$avgsteps, type="l", 
     xlab="5-minute interval", ylab="Avg. steps taken", main="Average daily activity pattern");
```
  
<br>
2. __Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?__  

To find this I use the _max()_ function in the average steps value and do subsetting looking for the corresponding interval.  

```{r}
avgStepsByInterval$interval[avgStepsByInterval$avgsteps == max(avgStepsByInterval$avgsteps)];
```
<br><br>


## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  

1. __Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)__  

I do this using the function _sum()_ for the steps whose value is NA.  

```{r}
sum(is.na(activityData$steps));
```
    
<br>
2. __Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.:__  

Taking a look at the data, we can observe that there are 288 different intervals for each date. Also, the NAs are all concentrated in the same eight days and when there is a NA in a specific day, all the values of the day are NAs. I decided to fill the missing values for this days with the average of each corresponding 5-minute interval for all the other days without NAs.  

```{r}
length(unique(activityData$interval));
table(activityData$date[is.na(activityData$steps)]);
```
  
<br>
3. __Create a new dataset that is equal to the original dataset but with the missing data filled in:__  

The process starts making a copy of the original data into _activityDataNAsFilled_. Then, a copy of the data with NAs is made in _activityDataNAs_. Using the _merge()_ function it's possible to add each average value to its corresponding data row. The common column for the merge is the interval. The result is merged into _activityDataNAsFilled_ and ordered using the _arrange()_ function. The NAs values in _steps_ are replaced by the average value and the extra column is deleted using _subset()_.  

```{r}
activityDataNAsFilled <- activityData;
activityDataNAs <- activityData[is.na(activityData$steps),];
activityDataNAs <- merge(activityDataNAs,avgStepsByInterval);
activityDataNAsFilled <- arrange(merge(activityDataNAsFilled,activityDataNAs, all = TRUE), date, interval);
activityDataNAsFilled$steps[is.na(activityDataNAsFilled$steps)] <- 
    activityDataNAsFilled$avgsteps[is.na(activityDataNAsFilled$steps)];
activityDataNAsFilled <- subset(activityDataNAsFilled, select = c("steps","date","interval"));
```
  
<br>
4. __Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?__  

From _activityDataNAsFilled_, I use the _aggregate()_ function to calculate the sum of the steps grouping by date. Then, I change the names of the columns into a more meaningful ones. The histogram is made with _hist()_ and _mean()_ and _median()_ functions are used.  

```{r}
sumStepsByDateNAsFilled <- aggregate(activityDataNAsFilled$steps, by=activityDataNAsFilled["date"], FUN=sum);
names(sumStepsByDateNAsFilled) <- c("date","sumsteps");
hist(sumStepsByDateNAsFilled$sumsteps, main="Histogram of the total number of steps by day (NAs filled)", 
     xlab="Total steps by day");
mean(sumStepsByDateNAsFilled$sumsteps);
median(sumStepsByDateNAsFilled$sumsteps);
```
  
  As I filled the NAs with the mean value for each interval, the mean takes the same value than the calculation made ignoring NAs. The median value is slightly higher and now it's the same as the mean. The frequencies of the histogram are raised, as some values are now more repeated.
<br><br>


## Are there differences in activity patterns between weekdays and weekends?
  
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.  

1. __Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.__  

This is performed using the _factor()_ function to encode the weekdays as an ordered factor. Then, the _levels()_ function is used to sort each day into two categories: "weekday" or "weekend".  

```{r}
activityDataNAsFilled$weekday <- weekdays(as.Date(activityDataNAsFilled$date));
activityDataNAsFilled$weekday <- factor(weekdays(as.Date(activityDataNAsFilled$date)), levels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), ordered = TRUE);
levels(activityDataNAsFilled$weekday) <- list(weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), weekend = c("Saturday", "Sunday"));
```
  
<br>
2. __Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.__  

The average by interval and weekday is made using _aggregate()_ and the names are changed again. The layout is divided in two setting _par(mfrow = ...)_ into two rows and one column. The newly created column _weekday_ is used to split the data between the two graphics. Finally, the plots are drawing using the Base plotting system plot function.

```{r}

avgStepsByIntervalNAsFilled <- aggregate(activityDataNAsFilled$steps, 
                                         by=activityDataNAsFilled[c("interval","weekday")], FUN=mean, na.rm = TRUE);
names(avgStepsByIntervalNAsFilled) <- c("interval","weekday","avgsteps");

par(mfrow = c(2,1));
with(avgStepsByIntervalNAsFilled, {
    plot(avgStepsByIntervalNAsFilled$interval[avgStepsByIntervalNAsFilled$weekday == "weekday"], 
         avgStepsByIntervalNAsFilled$avgsteps[avgStepsByIntervalNAsFilled$weekday == "weekday"], type="l", 
         xlab="5-minute interval", ylab="Avg. steps taken", main="Average weekday activity pattern");
    
    plot(avgStepsByIntervalNAsFilled$interval[avgStepsByIntervalNAsFilled$weekday == "weekend"], 
         avgStepsByIntervalNAsFilled$avgsteps[avgStepsByIntervalNAsFilled$weekday == "weekend"], type="l", 
         xlab="5-minute interval", ylab="Avg. steps taken", main="Average weekend activity pattern");
    
});
```  
  
  
The graphs evidence that in weekdays the steps are more concentrated in some intervals and in weekends the steps are more fairly distributed along the intervals.